Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-04 paper/metamonads.tex
continue with write up of section 4. add some comments in later sections too.

The use of a |functor| for making a container parametric is fairly 
straightforward.  It gets more interesting when one turns to the aspect
of what to return from the GE algorithm.

Functors and the generated return type: the Out structure.

@@

Show several examples (e.g., code with NullPivot and nodet: no traces
of either det nor pivoting). Then the code with RowPivoting, Det and
float domain.

@@

\section{Future work}\label{future}
connections with delimited continuations: making notation
more direct-style and potentially clearer.

@@

@@

>>>>

The use of a |functor| for making a container parametric is fairly 
straightforward.  It gets more interesting when one turns to the aspect
of what to return from the GE algorithm.  One could create an algebraic
data type (as was done in \cite{carette04}) to encode the various
choices; but this is wholly unsatisfying as we know that for any single
use, only one of the choices is ever possible, yet any routine which
calls the generated code must deal with these unreachable options.
Instead we use a module type with an \emph{abstract} type |res| for the
result type; the actual result type is then made by instantiating
this module type with different modules that decide what will be 
output: the matrix, the matrix and the rank, the matrix and the
determinant, the matrix, rank and determinant, and so on.  More
precisely, below we show this type and one instantiation.

\begin{small}
\begin{code}
module type OUTPUT = sig
  type contr
  type res
  module D : DETERMINANT
  module R : RANK
  module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list,
     ('a,'w) code) monad
end

module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) =
  struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det
  module R = Rank
  module P = DiscardPivot
  let make_result b = mdo {
    det  <-- D.fin ();
    rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. }
end
\end{code}
\end{small}
\noindent It is worthwhile noting that while the |OUTPUT| type is
relatively simple, instantiating it requires a rather complex functor
with some additional type constraints.  While we will explain some of
the details of the above code later, right now it suffices to notice
that |res| above is a 3-tuple with type |contr * Det.outdet * int|,
where in another instance this type is simply |contr|, and these are
formally incomparable types.

As is apparent from the output choices, several different quantities
\emph{may} need to be tracked in a particular GE implementation.
We therefore need to be able to conditionally generate code-variables,
and weave in corresponding tracking code for each of these quantities.
To be precise, we may need to keep track (independently) of the rank,
the determinant and the permutation list.  These pieces of data then
become part of the \emph{state} that is tracked by our monad.  To have
all this choice when needed, and yet have our code be modular and
composable as well as ensuring that the generated code does not 
contain any abstraction artifacts, it is important to make this 
state modular.  As this discussion is rather abstract, an example
will likely illustrate our point more clearly: we will show the 
code for tracking the determinant.

\begin{small}
\begin{code}
module type DETERMINANT = sig
  type indet
  type outdet
  type tdet = outdet ref
  type 'a lstate
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val zero_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val acc : ('a,indet) code -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val get : unit ->
    (('a,tdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val set : ('a,indet) code -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val fin : unit -> 
    (('a,outdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
end
\end{code}
\end{small}
\noindent  Determinant tracking is quite complex: we need to 
declare that we are tracking the determinant (|decl|), have some functions
for tracking the sign of the determinant (|upd_sign|,|zero_sign|),
|set|, |get|, accumulate (|acc|) unsigned values for the determinant, as 
well as returning a final value (|fin|) for the determinant.  In some cases,
like for float matrices where we do not return the determinant, all of
the above should generate no code at all, while for integer matrices where
we want the determinant as part of the output, each of these need to be
implemented.

\jacques{I am stopping my description of state here -- what I have
written above is quite dull, and not making the point very well.
I will need to revisit this}

@@

Show several examples (e.g., code with NullPivot and nodet: no traces
of either det nor pivoting). Then the code with RowPivoting, Det and
float domain. \jacques{which would really be great if we had a lot 
more space, but I think we're already long, so something is going to
have to be cut.  Maybe not this, but something!}

@@

\jacques{What do we consider \emph{related work} here?  Do we essentially gloss 
over related work in numerical software, code generation in general,
aspect-oriented stuff?  We don't want this section to be too long, but
not so short that we make it seem like we're inventing more than we
really are}

\section{Future work}\label{future}
connections with delimited continuations: making notation
more direct-style and potentially clearer.  More syntactic sugar
would help a lot here.

@@

Making all of the algorithmic code be completely independent of 
MetaOCaml markup and isolate all the code generation to a select
few 'base' modules and functors.  This should allow for simpler
debugging, where one can choose to instantiate either a direct-style
set of base modules for debugging, or a code-generation style set
of base modules for generating optimized code.

@@

Relation with aspect-oriented code: in AspectJ, aspects are (comparatively)
lightly typed, and are post-facto extensions of an existing piece of
code.  Here aspects are weaved together ``from scracth'' to make up a
piece of code/functionality.  One can understand previous work to be
more akin to dynamically typed aspect weaving, while we have started
investigating statically typed aspect weaving.
